---
title: "02_staar_2016_2017_cleaning"
author: "Tom"
date: "May 7, 2018"
output: html_document
---


## Goals

* Properly code NA values
* Convert from wide to tall format
* Assign descriptive labels to factors
* Split into datasets (approaching vs meeting)

```{r, echo = F, message=F, warning=F}
# load libraries
library(tidyverse)
library(readxl)
library(here)
library(writexl)
```

```{r}
# read in raw data
raw_df <- read_excel(here("raw_data/teach_trust.xlsx"))

# read in the group_labels from our "database" we made previously
group_labels <- read_excel(here("output_data/code_database.xlsx"))

raw_df # again notice there are negative values (-1 and -3) and "."
```


### Properly code NAs

```{r}
# negative students doesn't make sense to me
# and likely are a mistake or potentially a NA?
# "."" likely means missing data 

na_values <- c(".", "-1", "-3")

clean_df <- raw_df %>% 
    map_df(., ~replace(.x, .x %in% na_values, NA))

# sanity check data - Looks good there are now NAs instead of . or negative numbers
clean_df
```
## Tidy up the data

We need to generate reading scores for Meeting and Approaching Grade level. We will be excluding grade type == B as this is an unknown. We know E, M, and S. B could equal *both?* but excluding as unknown. I'll also split out the specific codes of interest rather than keeping them as one long code. Eg race/gender will be seperate from subject and year.

## Meeting grade level data

I wrote a function to take care of the vast majority of our data tidying. We'll create data subsets for the Numerator, Denominator, and Rate 

```{r}

# function does vast majority of our consistent tidying in a single chain
tall_function <- function(type, subject) {
    clean_df %>% 
    # select only the columns ending with type of interest (N/D/R)
    select(CAMPUS, GRDTYPE, ends_with(type)) %>%
    # lowercase these names
    rename(campus = CAMPUS, gradetype = GRDTYPE) %>%
    # gather the data from wide to tall
    gather(code, number_of_students, -campus, -gradetype) %>% 
    # exclude B grade and find classes that match subject of interest
    filter(str_detect(code, subject), gradetype != "B") %>%
    # since the code is consistent we can extract specific strings by location
    # from each of the codes of interest
    mutate(group_code = substr(code, start = 1, stop = 4),   
           subject_code = substr(code, start = 5, stop = 9), 
           year_code = paste0("20", substr(code, start = 10, stop = 11)), 
           type_code = substr(code, start = 12, stop = 12)) %>%
    mutate_at(vars(group_code, subject_code, year_code, type_code), str_trim) %>% 
    mutate(group_code = factor(group_code, levels = group_labels$group_code, labels = group_labels$group_labels),
           number_of_students = as.numeric(number_of_students),
           year_code = factor(year_code),
           gradetype = factor(gradetype, levels = c("E", "M", "S"), labels = c("Elementary", "Middle-school", "Secondary")))
}

# meeting grade level denominator
meet_denom <- tall_function("D", "AR042") 

# meeting grade level numerator
meet_num <- tall_function("N", "AR042") 

# meeting grade level rate
meet_rate <- tall_function("R", "AR042")

# sanity check the dimensions
list(meet_denom, meet_num, meet_rate) %>% 
    map(., dim)

# Dimensions look good -- should be 6664 x 8, original raw_df was 6664
```

## Join the meeting dataframes into one master

Using `dplyr::inner_joins` we will merge the numerator, denominator, and rate dataframes to compare and generate a master dataframe. For some reason I am still seeing negative values for some districts, so we will filtering only students that are > 0. I don't see a situation where you can have negative students in the denominator (maybe leaving school?) but regardless I am making the assumption it is a clerical error and will be excluding it.

```{r}

meeting_df <- inner_join(meet_num, meet_denom, by = c("campus", "gradetype", "year_code", "group_code", "subject_code")) %>% 
    rename(numer_students = number_of_students.x, denom_students = number_of_students.y) %>% 
    select(-ends_with(".y"), - ends_with(".x")) %>% 
    inner_join(meet_rate, by = c("campus", "gradetype", "year_code", "group_code", "subject_code")) %>% 
    filter(numer_students > 0, denom_students > 0) %>% 
    mutate(rate = numer_students/denom_students,
           given_rate = number_of_students/100) %>% 
    select(-number_of_students)

 
# sanity check the rate vs given rate from the dataset
meeting_df %>% 
    mutate(sanity_check = round(rate - given_rate, digits = 3)) %>%
    ggplot(aes(x = sanity_check)) +
    geom_histogram() +
    labs(title = "~ -0.005% to 0.005% variation between my calculated rate and the actual rate")

# I'll take that as rate being == percentage!

```

```{r}
# sanity check overall data w/ a basic graph
# I'm probably going to exclude pacific islander and American Indian from the final graph  
# as there is substantial missing data or just no students in these racial categories
meet_rate %>% 
  group_by(group_code, gradetype, year_code) %>% 
  summarize(mean = mean(number_of_students, na.rm = T)) %>% 
  ungroup %>% 
  ggplot(aes(x = forcats::fct_reorder(group_code, mean), y = mean, group = group_code, color = year_code)) +
  geom_point() +
  coord_flip() +
  facet_grid(~gradetype)
 
meet_rate %>% 
  group_by(group_code, gradetype, year_code) %>% 
  summarize(mean = mean(number_of_students, na.rm = T)) %>% 
  ungroup %>% 
    filter(is.na(mean))
```

## Approaching grade level data

I am using our previously defined function to take care of the vast majority of our data tidying. We'll create data subsets for the Numerator, Denominator, and Rate. 

```{r}

# meeting grade level denominator
approach_denom <- tall_function("D", "AR01") 

# meeting grade level numerator
approach_num <- tall_function("N", "AR01") 

# meeting grade level rate
approach_rate <- tall_function("R", "AR01")

# sanity check the dimensions
list(approach_denom, approach_num, approach_rate) %>% 
    map(., dim)

# Dimensions look good -- should be 6664 x 8, original raw_df was 6664
```

## Join the approach dataframes into one master

Using `dplyr::inner_joins` we will merge the numerator, denominator, and rate dataframes to compare and generate a master dataframe. For some reason I am still seeing negative values for some districts, so we will filtering only students that are > 0. I don't see a situation where you can have negative students in the denominator (maybe leaving school?) but regardless I am making the assumption it is a clerical error and will be excluding it. Importantly, the `subject_code` does not match between 2016 and 2017 but as far as I can tell it represents the same thing (reading) but regardless we will not join on it as it will prevent them from joining at all.

```{r}
# join the dataframe, filter students > 0, and calculate rate
approach_df <- inner_join(approach_num, approach_denom, by = c("campus", "gradetype", "year_code", "group_code")) %>% 
    rename(numer_students = number_of_students.x, denom_students = number_of_students.y) %>% 
    select(-ends_with(".y"), - ends_with(".x")) %>% 
    inner_join(approach_rate, by = c("campus", "gradetype", "year_code", "group_code")) %>% 
    filter(numer_students > 0, denom_students > 0) %>% 
    mutate(rate = numer_students/denom_students,
           given_rate = number_of_students/100) %>% 
    select(-number_of_students)

 
# sanity check the rate vs given rate from the dataset
approach_df %>% 
    mutate(sanity_check = round(rate - given_rate, digits = 3)) %>%
    arrange(desc(sanity_check))

```

```{r}
# sanity check overall data w/ a basic graph
# I'm probably going to exclude pacific islander and American Indian from the final graph  
# as there is substantial missing data or just no students in these racial categories
approach_rate %>% 
  group_by(group_code, gradetype, year_code) %>% 
  summarize(mean = mean(number_of_students, na.rm = T)) %>% 
  ungroup %>% 
  ggplot(aes(x = forcats::fct_reorder(group_code, mean), y = mean, group = group_code, color = year_code)) +
  geom_point() +
  coord_flip() +
  facet_grid(~gradetype)
 

approach_rate %>% 
  group_by(group_code, gradetype, year_code) %>% 
  summarize(mean = mean(number_of_students, na.rm = T)) %>% 
  ungroup %>% 
    filter(is.na(mean))
```

### Save the clean data

We will save this clean data as both RDS and as excel file.

```{r}
saveRDS(approach_df, here("/output_data/approach_df.rds"))
saveRDS(meeting_df, here("/output_data/meeting_df.rds"))

list(approaching = approach_df, meeting = meeting_df) %>% 
    writexl::write_xlsx(here("/output_data/clean_df.xlsx"))
```

